---
title: "Flight Delay Prediction Model "
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: 
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  cache.lazy = FALSE
)
```

# Executive Summary

**Objective**: Develop machine learning models to predict flight delays using historical operational and meteorological data. And possibly use arrival classification model to analyze when connection insurance is worth it for travelers.

**Dataset**: 624,855 U.S. domestic flights (2010-2019) with temporal, operational, and weather features.

---

# 1. Data Preparation {.tabset}
If needed these will instal all packages
```{r install-packages, eval=FALSE}
install.packages(c(
  "arrow", "dplyr", "xgboost", "pROC", "caret", "lubridate",
  "zoo", "ggplot2", "tidyr", "knitr", "stringr", "purrr",
  "kableExtra", "rBayesianOptimization"
))
```

## 1.1 Data Cleaning

```{r data-cleaning}
library(arrow)
library(dplyr)
library(xgboost)
library(pROC)
library(caret)
library(lubridate)
library(zoo)
library(ggplot2)
library(tidyr)
library(knitr)
library(stringr)
library(purrr)
library(kableExtra)
library(rBayesianOptimization)

set.seed(42)

# Define booking features (EXACTLY same as cancellation)
booking_features <- c(
  "Year", "Quarter", "Month", "DayofMonth", "DayOfWeek",
  "CRSDepTime", "CRSElapsedTime", 
  "Reporting_Airline", "Origin", "Dest", 
  "Distance", "OriginWac", "DestWac", "avg_fare",
  "origin_TEMP", "origin_DEWP", "origin_VISIB", "origin_WDSP", "origin_MXSPD", 
  "origin_MAX", "origin_MIN", "origin_PRCP", "origin_SNDP",
  "dest_TEMP", "dest_DEWP", "dest_VISIB", "dest_WDSP", "dest_MXSPD",
  "dest_MAX", "dest_MIN", "dest_PRCP", "dest_SNDP"
)

# Load data 
df_full <- read_parquet("merged_weather_final.parquet")

airline_levels <- sort(unique(df_full$Reporting_Airline))
origin_levels <- sort(unique(df_full$Origin))
dest_levels <- sort(unique(df_full$Dest))

df_full <- df_full %>%
  mutate(
    Reporting_Airline = as.integer(factor(Reporting_Airline, levels = airline_levels)),
    Origin = as.integer(factor(Origin, levels = origin_levels)),
    Dest = as.integer(factor(Dest, levels = dest_levels))
  )

# Split data 
train_data <- df_full %>% filter(Year != 2019)
test_data <- df_full %>% filter(Year == 2019)

encode_and_clean <- function(data) {
  data %>% 
    select(where(~ !is.character(.)))
}

```

ARRIVAL DELAY INSURANCE: DATA PREPARATION


Dataset statistics:
  Total flights: 635,311 
  Delayed ≥15 min: 117,795 (18.90%)
  On-time: 505,549 

Encoding categorical variables...
✓ Encoded 22 airlines
✓ Encoded 369 origins
✓ Encoded 368 destinations

Train/test split:
  Train (2010-2018): 559,075 
  Test (2019): 76,236 
## 1.2 Feature Engineering

```{r features}
# Prepare datasets
dep_train <- train_data %>%
  filter(!is.na(DepDel15), !is.na(DepDelay)) %>%
  select(all_of(c(booking_features, "DepDel15", "DepDelay"))) %>%
  encode_and_clean()

dep_test <- test_data %>%
  filter(!is.na(DepDel15), !is.na(DepDelay)) %>%
  select(all_of(c(booking_features, "DepDel15", "DepDelay"))) %>%
  encode_and_clean()

arr_train <- train_data %>%
  filter(!is.na(ArrDel15), !is.na(ArrDelay)) %>%
  select(all_of(c(booking_features, "ArrDel15", "ArrDelay"))) %>%
  encode_and_clean()

arr_test <- test_data %>%
  filter(!is.na(ArrDel15), !is.na(ArrDelay)) %>%
  select(all_of(c(booking_features, "ArrDel15", "ArrDelay"))) %>%
  encode_and_clean()

# Overall delay statistics
delayed_flights <- train_data %>% filter(DepDelay > 15, !is.na(DepDelay))

```

Delay Distribution Analysis

Total flights: 559075 
Delayed flights: 96871 
Delay rate: 17.3 %

Delay Severity Breakdown:
  0-15 min (on-time):   449669 (81.7%) 
  15-30 min (minor):    37894 (6.9%) 
  30-60 min (moderate): 30903 (5.6%) 
  60+ min (major):      31813 (5.8%) 
---

# 2. Model Development: Comparing Three XGBoost Approaches {.tabset}

## 2.1 Three-Way Comparison: Baseline vs Calibration vs Ranking

**Question:** Which XGBoost approach works best for our 18% delay rate: baseline, calibration-focused (max_delta_step), or ranking-focused (scale_pos_weight)? The official xgboost website recommends using either calibration for optimal precision, or max delta step for optimal ranking model.

```{r three-approaches-comparison}
set.seed(42)
# 1. DEFINE TRAINING & EVALUATION FUNCTIONS---------------------------------------------------------------

train_baseline <- function(train_df, test_df, target) {
  dtrain <- xgb.DMatrix(as.matrix(train_df %>% select(-any_of(c("DepDel15", "DepDelay", "DepDel30", "DelT", "ArrDel15", "ArrDelay", "ArrDel30", "ArrDel60", "ArrDel90", "ArrDel120", "Year")))), label = train_df[[target]])
  dtest <- xgb.DMatrix(as.matrix(test_df %>% select(-any_of(c("DepDel15", "DepDelay", "DepDel30", "DelT", "ArrDel15", "ArrDelay", "ArrDel30", "ArrDel60", "ArrDel90", "ArrDel120", "Year")))), label = test_df[[target]])
  
  params <- list(objective = "binary:logistic", eval_metric = "logloss", max_depth = 4, eta = 0.05,
                 min_child_weight = 5, subsample = 0.8, colsample_bytree = 0.8)
  
  model <- xgb.train(params, dtrain, nrounds = 200, watchlist = list(test = dtest), early_stopping_rounds = 20, verbose = 0)
  list(model = model, dtrain = dtrain, dtest = dtest)
}

train_calibrated <- function(train_df, test_df, target) {
  dtrain <- xgb.DMatrix(as.matrix(train_df %>% select(-any_of(c("DepDel15", "DepDelay", "DepDel30", "DelT", "ArrDel15", "ArrDelay", "ArrDel30", "ArrDel60", "ArrDel90", "ArrDel120", "Year")))), label = train_df[[target]])
  dtest <- xgb.DMatrix(as.matrix(test_df %>% select(-any_of(c("DepDel15", "DepDelay", "DepDel30", "DelT", "ArrDel15", "ArrDelay", "ArrDel30", "ArrDel60", "ArrDel90", "ArrDel120","Year")))), label = test_df[[target]])
  
  params <- list(objective = "binary:logistic", eval_metric = "logloss", max_depth = 4, eta = 0.05,
                 min_child_weight = 5, subsample = 0.8, colsample_bytree = 0.8, max_delta_step = 1)
  
  model <- xgb.train(params, dtrain, nrounds = 200, watchlist = list(test = dtest), early_stopping_rounds = 20, verbose = 0)
  list(model = model, dtrain = dtrain, dtest = dtest)
}

train_ranking <- function(train_df, test_df, target) {
  dtrain <- xgb.DMatrix(as.matrix(train_df %>% select(-any_of(c("DepDel15", "DepDelay", "DepDel30", "DelT", "ArrDel15", "ArrDelay", "ArrDel30", "ArrDel60", "ArrDel90", "ArrDel120","Year")))), label = train_df[[target]])
  dtest <- xgb.DMatrix(as.matrix(test_df %>% select(-any_of(c("DepDel15", "DepDelay", "DepDel30", "DelT", "ArrDel15", "ArrDelay", "ArrDel30", "ArrDel60", "ArrDel90", "ArrDel120","Year")))), label = test_df[[target]])
  
  neg <- sum(train_df[[target]] == 0); pos <- sum(train_df[[target]] == 1)
  params <- list(objective = "binary:logistic", eval_metric = "auc", max_depth = 4, eta = 0.05,
                 min_child_weight = 5, subsample = 0.8, colsample_bytree = 0.8, scale_pos_weight = neg/pos)
  
  model <- xgb.train(params, dtrain, nrounds = 200, watchlist = list(test = dtest), early_stopping_rounds = 20, verbose = 0)
  list(model = model, dtrain = dtrain, dtest = dtest)
}

evaluate_model <- function(model_result, test_df, target, approach_name, threshold_label) {
  preds <- predict(model_result$model, model_result$dtest)
  roc_obj <- roc(test_df[[target]], preds, quiet = TRUE)
  
  # Safely extract exactly ONE numeric threshold, even if there are ties
  best_coords <- coords(roc_obj, "best", ret = "threshold", transpose = FALSE)
  
  # best_coords could be a data.frame or vector depending on pROC version/ties.
  # Extracting [1, 1] and forcing to numeric ensures we only get a single scalar value.
  best_thresh <- as.numeric(best_coords[1, 1]) 
  
  # Generate predicted classes using the single threshold
  predicted_classes <- as.factor(ifelse(preds > best_thresh, 1, 0))
  actual_classes <- as.factor(test_df[[target]])
  
  # Build confusion matrix
  cm <- confusionMatrix(predicted_classes, actual_classes, positive = "1")
  
  precision <- cm$byClass["Pos Pred Value"]
  recall <- cm$byClass["Sensitivity"]
  f1 <- 2 * (precision * recall) / (precision + recall)
  
  data.frame(
    Threshold = threshold_label,
    Approach = approach_name,
    AUC = round(auc(roc_obj), 4),
    Kappa = round(cm$overall["Kappa"], 3),
    Precision = sprintf("%.1f%%", cm$byClass["Pos Pred Value"] * 100),
    Recall = sprintf("%.1f%%", cm$byClass["Sensitivity"] * 100),
    Specificity = sprintf("%.1f%%", cm$byClass["Specificity"] * 100),
    `F1-Score` = round(f1, 3),
    check.names = FALSE
  )
}

# 2. RUN DEPARTURE DELAY MODELS (≥15, ≥30)----------------------------------------------------------------
dep_baseline_15 <- train_baseline(dep_train, dep_test, "DepDel15")
dep_calib_15 <- train_calibrated(dep_train, dep_test, "DepDel15")
dep_rank_15 <- train_ranking(dep_train, dep_test, "DepDel15")

results_15 <- rbind(
  evaluate_model(dep_baseline_15, dep_test, "DepDel15", "Baseline (no handling)", "Dep ≥15 min"),
  evaluate_model(dep_calib_15, dep_test, "DepDel15", "Calibration (max_delta_step)", "Dep ≥15 min"),
  evaluate_model(dep_rank_15, dep_test, "DepDel15", "Ranking (scale_pos_weight)", "Dep ≥15 min")
)

dep_train_30 <- dep_train %>% mutate(DepDel30 = ifelse(DepDelay >= 30, 1, 0))
dep_test_30 <- dep_test %>% mutate(DepDel30 = ifelse(DepDelay >= 30, 1, 0))

dep_baseline_30 <- train_baseline(dep_train_30, dep_test_30, "DepDel30")
dep_calib_30 <- train_calibrated(dep_train_30, dep_test_30, "DepDel30")
dep_rank_30 <- train_ranking(dep_train_30, dep_test_30, "DepDel30")

results_30 <- rbind(
  evaluate_model(dep_baseline_30, dep_test_30, "DepDel30", "Baseline (no handling)", "Dep ≥30 min"),
  evaluate_model(dep_calib_30, dep_test_30, "DepDel30", "Calibration (max_delta_step)", "Dep ≥30 min"),
  evaluate_model(dep_rank_30, dep_test_30, "DepDel30", "Ranking (scale_pos_weight)", "Dep ≥30 min")
)

# 3. RUN ARRIVAL DELAY MODELS (≥15, ≥30, ≥60)-------------------------------------------------------------

# Train & Eval ≥15 min Arrival
arr_baseline_15 <- train_baseline(arr_train, arr_test, "ArrDel15")
arr_calib_15    <- train_calibrated(arr_train, arr_test, "ArrDel15")
arr_rank_15     <- train_ranking(arr_train, arr_test, "ArrDel15")

results_arr_15 <- rbind(
  evaluate_model(arr_baseline_15, arr_test, "ArrDel15", "Baseline (no handling)", "Arr ≥15 min"),
  evaluate_model(arr_calib_15,    arr_test, "ArrDel15", "Calibration (max_delta_step)", "Arr ≥15 min"),
  evaluate_model(arr_rank_15,     arr_test, "ArrDel15", "Ranking (scale_pos_weight)", "Arr ≥15 min")
)

# Train & Eval >=30 and >=60 min Arrivals — ranking only--------------------------------------------------
arr_extended_results_list <- list()

for (th in c(30, 60)) {
  target_name <- paste0("ArrDel", th)
  label_name  <- paste0("Arr >=", th, " min")
  
  tmp_train <- arr_train
  tmp_test  <- arr_test
  tmp_train[[target_name]] <- ifelse(tmp_train$ArrDelay >= th, 1, 0)
  tmp_test[[target_name]]  <- ifelse(tmp_test$ArrDelay  >= th, 1, 0)
  
  r_mod <- train_ranking(tmp_train, tmp_test, target_name)
  assign(paste0("arr_", th, "_class"), r_mod)
  
  arr_extended_results_list[[as.character(th)]] <- evaluate_model(
    r_mod, tmp_test, target_name, "Ranking (scale_pos_weight)", label_name
  )
}

results_arr_extended <- do.call(rbind, arr_extended_results_list)

# 4. COMBINED RESULTS-------------------------------------------------------------------------------------

all_results <- rbind(results_15, results_30, results_arr_15, results_arr_extended)
kable(all_results, caption = "Three Approaches Compared: Departure & Arrival Delay Thresholds", row.names = FALSE)

# 5. COMPARISON ANALYSIS (Saving Top Models)--------------------------------------------------------------

dep_class    <- dep_rank_15
dep_30_class <- dep_rank_30
arr_class    <- arr_rank_15
arr_30_class <- get("arr_30_class")
arr_60_class <- get("arr_60_class")

# Save DMatrices
dtest_dep    <- dep_class$dtest
dtest_arr    <- arr_class$dtest
dtest_arr_30 <- arr_30_class$dtest

```

**Key Findings:**
All approaches show similarly poor calibration (Kappa ~0.15–0.20) with only marginal performance differences. At ≥30 min, the task becomes harder due to lower prevalence—AUC improves slightly (better ranking), but precision drops.

Decision: Use the ranking approach (scale_pos_weight) since differences are minimal and our priority is identifying high-risk cases, aligning with XGBoost recommendations.

# 3. Classification vs Regression Performance Gap {.tabset}

## 3.1 Overview

A critical finding emerged during model development: while classification models achieved fair performance for binary delay prediction, regression models consistently failed to predict delay magnitude despite extensive exploration of model variants and feature engineering approaches.

This section documents the investigation into this performance gap and its implications.

## 3.2 Regression Performance Across Multiple Approaches

We now train regression models for for only positive delays, between 0 to 60 minutes, and a model including early departures as well(negative delays)
```{r regression-exploration, fig.height=5}
## Regression Exploration: Why Magnitude Prediction Fails

# Train regression model
train_regression <- function(train_df, test_df, target, filter_fn = NULL) {
  if (!is.null(filter_fn)) {
    train_df <- filter_fn(train_df)
    test_df <- filter_fn(test_df)
  }
  
  dtrain <- xgb.DMatrix(as.matrix(train_df %>% select(-any_of(c("DepDel15", "DepDelay", "DepDel30", "DelT", "ArrDel15", "ArrDelay")))), label = train_df[[target]])
  dtest <- xgb.DMatrix(as.matrix(test_df %>% select(-any_of(c("DepDel15", "DepDelay", "DepDel30", "DelT", "ArrDel15", "ArrDelay")))), label = test_df[[target]])
  
  params <- list(objective = "reg:squarederror", eval_metric = "rmse", max_depth = 4, eta = 0.05,
                 min_child_weight = 5, subsample = 0.8, colsample_bytree = 0.8)
  
  model <- xgb.train(params, dtrain, nrounds = 200, watchlist = list(test = dtest), early_stopping_rounds = 20, verbose = 0)
  preds <- predict(model, dtest)
  actual <- getinfo(dtest, "label")
  
  list(preds = preds, actual = actual, r = cor(preds, actual), pred_sd = sd(preds), actual_sd = sd(actual), n = length(actual))
}

# Test approaches
approaches <- list(
  list(name = "All delays", filter_fn = NULL, range = "(-42, 1855)"),
  list(name = "Positive only", filter_fn = function(df) df %>% filter(DepDelay >= 0), range = "(0, 1855)"),
  list(name = "Normal (0-60 min)", filter_fn = function(df) df %>% filter(DepDelay >= 0, DepDelay <= 60), range = "(0, 60)")
)

results <- lapply(approaches, function(a) c(list(name = a$name, range = a$range), train_regression(dep_train, dep_test, "DepDelay", a$filter_fn)))

# Classification comparison
dtrain_class <- xgb.DMatrix(as.matrix(dep_train %>% select(-DepDel15, -DepDelay)), label = dep_train$DepDel15)
dtest_class <- xgb.DMatrix(as.matrix(dep_test %>% select(-DepDel15, -DepDelay)), label = dep_test$DepDel15)
model_class <- xgb.train(params = list(objective = "binary:logistic", eval_metric = "auc", max_depth = 4, eta = 0.05,
                                       min_child_weight = 5, subsample = 0.8, colsample_bytree = 0.8, 
                                       scale_pos_weight = sum(dep_train$DepDel15 == 0)/sum(dep_train$DepDel15 == 1)),
                         data = dtrain_class, nrounds = 200, verbose = 0)
class_roc <- roc(dep_test$DepDel15, predict(model_class, dtest_class), quiet = TRUE)

# Table
comparison <- data.frame(
  Approach = c(sapply(results, function(x) x$name), "Classification"),
  Range = c(sapply(results, function(x) x$range), "Binary (≥15 min)"),
  N = c(sapply(results, function(x) x$n), nrow(dep_test)),
  Performance = c(sapply(results, function(x) sprintf("r = %.3f", x$r)), sprintf("AUC = %.3f", auc(class_roc))),
  Pred_SD = c(sapply(results, function(x) sprintf("%.1f", x$pred_sd)), "-"),
  Actual_SD = c(sapply(results, function(x) sprintf("%.1f", x$actual_sd)), "-")
)
kable(comparison, caption = "Regression vs Classification Performance", row.names = FALSE)

# Plots
plot_list <- lapply(seq_along(results), function(i) {
  r <- results[[i]]
  lims <- c(if(i==3) 0 else if(i==2) 0 else -20, if(i==3) 60 else 100)
  ggplot(data.frame(actual=r$actual, pred=r$preds), aes(actual, pred)) +
    geom_point(alpha=0.3, size=0.5) + geom_abline(color="red") + 
    geom_hline(yintercept=mean(r$preds), color="blue", linetype="dashed") +
    coord_cartesian(xlim=lims, ylim=lims) + theme_minimal() +
    labs(title=sprintf("%s (r = %.3f)", r$name, r$r), x="Actual Delay (min)", y="Predicted Delay (min)")
})
gridExtra::grid.arrange(grobs=plot_list, ncol=2)

# Variance
var_data <- do.call(rbind, lapply(results, function(r) data.frame(Approach=r$name, Type=c("Predicted","Actual"), SD=c(r$pred_sd, r$actual_sd))))
ggplot(var_data, aes(Approach, SD, fill=Type)) + geom_col(position="dodge") + 
  scale_fill_manual(values=c("steelblue","coral")) + theme_minimal() + theme(legend.position="top") +
  labs(title="Predicted vs Actual Variance", x=NULL, y="SD (minutes)")
```

**Key Finding:** Regression predictions show extremely low variance (SD 2-3 min) vs actual (SD 15-72 min). Model defaults to near-constant predictions regardless of flight conditions, indicating inability to differentiate delay magnitudes. Classification succeeds (AUC 0.69) where regression fails. This is possibly due to that the variables in our dataset can predict whether delay can happen but not the amount this delay will be, which might depend on factors not included in our dataset.





## 3.3 Cascade Modeling: Departure Probabilities as Arrival Features {.tabset}

### 3.3.1 Motivation

Given the high correlation between departure and arrival delays (r = 0.95), we tested whether departure delay probabilities could enhance arrival predictions through cascade modeling. The hypothesis: departure probability predictions (P(DepDel≥15), P(DepDel≥30), etc.) might encode operational risk patterns beyond raw weather features alone.

### 3.3.2 Implementation

```{r cascade-modeling, cache=TRUE}
# Create matched datasets
train_matched <- train_data %>%
  filter(!is.na(DepDel15), !is.na(DepDelay), !is.na(ArrDel15), !is.na(ArrDelay)) %>%
  select(all_of(c(booking_features, "DepDel15", "DepDelay", "ArrDel15", "ArrDelay", "Year"))) %>%
  encode_and_clean()

test_matched <- test_data %>%
  filter(!is.na(DepDel15), !is.na(DepDelay), !is.na(ArrDel15), !is.na(ArrDelay)) %>%
  select(all_of(c(booking_features, "DepDel15", "DepDelay", "ArrDel15", "ArrDelay", "Year"))) %>%
  encode_and_clean()

# Use our two trained threshold models: 15 and 30 min
thresholds <- c(15, 30)

set.seed(123)
folds <- createFolds(train_matched$DepDel15, k = 5)
cv_dep_probs <- matrix(0, nrow(train_matched), length(thresholds))

for (i in seq_along(thresholds)) {
  cv_preds <- numeric(nrow(train_matched))
  for (fold_idx in 1:5) {
    val_idx <- folds[[fold_idx]]
    fold_train <- train_matched[-val_idx, ] %>% 
      mutate(DelT = ifelse(DepDelay >= thresholds[i], 1, 0))
    fold_val <- train_matched[val_idx, ] %>%
      mutate(DelT = ifelse(DepDelay >= thresholds[i], 1, 0))
    
    # Train with ranking optimization (consistent with our final approach)
    fold_result <- train_ranking(fold_train, fold_val, "DelT")
    
    # FIX 1: Added -any_of("Year") to drop the 32nd column
    cv_preds[val_idx] <- predict(fold_result$model, xgb.DMatrix(
      as.matrix(fold_val %>% select(-DepDel15, -DepDelay, -ArrDel15, -ArrDelay, -DelT, -any_of("Year")))
    ))
  }
  # FIX 2: Removed the rogue/duplicate predict call that was sitting right here
}

# FIX 3: Added -any_of("Year") to both test prediction calls below
test_dep_probs <- cbind(
  predict(dep_class$model, xgb.DMatrix(as.matrix(test_matched %>% select(-DepDel15, -DepDelay, -ArrDel15, -ArrDelay, -any_of("Year"))))),
  predict(dep_30_class$model, xgb.DMatrix(as.matrix(test_matched %>% select(-DepDel15, -DepDelay, -ArrDel15, -ArrDelay, -any_of("Year")))))
)

# Create datasets
train_base <- train_matched %>% select(-DepDel15, -DepDelay)
test_base <- test_matched %>% select(-DepDel15, -DepDelay)
train_enh <- train_base
test_enh <- test_base

# Add departure probability features
for (i in seq_along(thresholds)) {
  train_enh[[paste0("P_Dep", thresholds[i])]] <- cv_dep_probs[, i]
  test_enh[[paste0("P_Dep", thresholds[i])]] <- test_dep_probs[, i]
}

# Train baseline model (weather features only, no departure probabilities)
neg_base <- sum(train_base$ArrDel15 == 0)
pos_base <- sum(train_base$ArrDel15 == 1)

model_base <- xgb.train(
  params = list(
    objective = "binary:logistic", 
    eval_metric = "auc",
    eta = 0.05, max_depth = 4,
    subsample = 0.8, colsample_bytree = 0.8,
    scale_pos_weight = neg_base/pos_base
  ),
  data = xgb.DMatrix(as.matrix(train_base %>% select(-ArrDel15, -ArrDelay)),
                     label = train_base$ArrDel15),
  nrounds = 200, verbose = 0
)

# Train enhanced model (weather features + departure probabilities)
neg_enh <- sum(train_enh$ArrDel15 == 0)
pos_enh <- sum(train_enh$ArrDel15 == 1)

model_enh <- xgb.train(
  params = list(
    objective = "binary:logistic",
    eval_metric = "auc",
    eta = 0.05, max_depth = 4,
    subsample = 0.8, colsample_bytree = 0.8,
    scale_pos_weight = neg_enh/pos_enh
  ),
  data = xgb.DMatrix(as.matrix(train_enh %>% select(-ArrDel15, -ArrDelay)),
                     label = train_enh$ArrDel15),
  nrounds = 200, verbose = 0
)

# Evaluate both models
preds_base <- predict(model_base, xgb.DMatrix(as.matrix(test_base %>% select(-ArrDel15, -ArrDelay))))
preds_enh <- predict(model_enh, xgb.DMatrix(as.matrix(test_enh %>% select(-ArrDel15, -ArrDelay))))
roc_base <- roc(test_base$ArrDel15, preds_base, quiet = TRUE)
roc_enh <- roc(test_enh$ArrDel15, preds_enh, quiet = TRUE)

cascade_comparison <- data.frame(
  Model = c("Baseline (weather only)", "Enhanced (+ departure probabilities)"),
  Features = c("31 booking-time features", "31 booking-time + 2 departure probs"),
  AUC = c(round(auc(roc_base), 4), round(auc(roc_enh), 4)),
  `AUC Gain` = c("-", sprintf("+%.4f", auc(roc_enh) - auc(roc_base))),
  check.names = FALSE
)

kable(cascade_comparison, caption = "Cascade Modeling: Departure Probabilities as Features", row.names = FALSE)
```

### 3.3.3 Results

```{r cascade-analysis}
# Statistics for plot
improvement <- auc(roc_enh) - auc(roc_base)
pct_imp <- 100 * improvement / auc(roc_base)
imp_enh <- xgb.importance(model_enh)
dep_feats <- imp_enh[grepl("^P_Dep", imp_enh$Feature), ]
print(dep_feats, row.names = FALSE)
# Plot
par(mfrow = c(1, 2))
plot(roc_base, col = "steelblue", lwd = 2, main = "ROC Comparison")
plot(roc_enh, col = "coral", lwd = 2, add = TRUE)
legend("bottomright", c(sprintf("Base %.4f", auc(roc_base)),
                        sprintf("Enh %.4f", auc(roc_enh))),
       col = c("steelblue", "coral"), lwd = 2, cex = 0.8)
top12 <- head(imp_enh, 12)
barplot(top12$Gain, names.arg = top12$Feature, las = 2, cex.names = 0.6,
        col = ifelse(grepl("^P_Dep", top12$Feature), "coral", "steelblue"),
        main = "Feature Importance", ylab = "Gain")
par(mfrow = c(1, 1))
```
The cascade model, which incorporates departure delay probabilities (P_Dep15, P_Dep30) as additional features, achieves only a marginal improvement over the baseline weather-only model (AUC: 0.6921 → 0.6955, +0.34%). Despite the strong correlation between departure and arrival delays (r = 0.96), the departure probability features dominate the feature importance ranking but provide minimal predictive lift. This suggests that departure probabilities largely encode the same weather and operational patterns already captured by the baseline features, rather than contributing novel information. The result validates using separate, independent models for departure and arrival prediction—cascade modeling adds computational complexity without meaningful performance gains.

# 5. Arrival Delay Insurance: Booking-Time Prediction {.tabset}

While a live weather delay model can be useful, it is not representative for a booking time model. If we want customers to see their delay risk at booking time, we need a model trained on **historical weather patterns**. This section trains **two approaches**: pure monthly averages (baseline) and hybrid monthly + 7-day smoothed visibility (enhanced), then compares their performance. We use arrival delay for this, since this is most usefull at booking time especially when booking multiple flights, as this can mean you might miss your connection.

## 5.1 Prepare Data with ≥15 Minute Target
```{r prepare-delay-data}
# Data Preparation =============================================================

# Prepare data with ≥15 minute delay target
delay_data <- train_data %>%  
  filter(!is.na(ArrDelay)) %>%
  mutate(
    Delayed_15 = ifelse(ArrDelay >= 15, 1, 0),
    day_of_year = yday(as.Date(paste(Year, Month, DayofMonth, sep = "-")))
  ) %>%
  select(all_of(booking_features), Delayed_15, day_of_year)

# Preserve day_of_year before encoding
day_of_year_vec <- delay_data$day_of_year
delay_data <- encode_and_clean(delay_data)
delay_data$day_of_year <- day_of_year_vec

delay_train <- delay_data %>% filter(Year <= 2018)

# Test Set Preparation =========================================================

delay_test_raw <- test_data %>%
  filter(!is.na(ArrDelay)) %>%
  mutate(
    Delayed_15 = ifelse(ArrDelay >= 15, 1, 0),
    day_of_year = yday(as.Date(paste(Year, Month, DayofMonth, sep = "-")))
  ) %>%
  select(all_of(booking_features), Delayed_15, day_of_year, ArrDelay)

day_of_year_vec_test <- delay_test_raw$day_of_year
arr_delay_vec_test   <- delay_test_raw$ArrDelay
delay_test_raw <- encode_and_clean(delay_test_raw)
delay_test_raw$day_of_year <- day_of_year_vec_test
delay_test_raw$ArrDelay    <- arr_delay_vec_test

delay_test <- delay_test_raw
```

## 5.2 Compute Historical Weather (Both Approaches)
```{r delay-weather-computation}
# Approach 1: Pure Monthly Averages ============================================

monthly_all_origin <- delay_train %>%
  group_by(Origin, Month) %>%
  summarise(
    across(starts_with("origin_"), ~ mean(., na.rm = TRUE)),
    .groups = "drop"
  )

monthly_all_dest <- delay_train %>%
  group_by(Dest, Month) %>%
  summarise(
    across(starts_with("dest_"), ~ mean(., na.rm = TRUE)),
    .groups = "drop"
  )

# Approach 2: Hybrid (Monthly + 7-Day VISIB) ===================================

# Monthly averages for non-visibility features
monthly_nonvisib_origin <- delay_train %>%
  group_by(Origin, Month) %>%
  summarise(
    across(starts_with("origin_") & !contains("VISIB"), ~ mean(., na.rm = TRUE)),
    .groups = "drop"
  )

monthly_nonvisib_dest <- delay_train %>%
  group_by(Dest, Month) %>%
  summarise(
    across(starts_with("dest_") & !contains("VISIB"), ~ mean(., na.rm = TRUE)),
    .groups = "drop"
  )

# 7-day smoothed visibility
origin_visib_7day <- delay_train %>%
  arrange(Origin, day_of_year) %>%
  group_by(Origin) %>%
  mutate(
    origin_VISIB_7day = zoo::rollmean(origin_VISIB, k = 7, fill = NA, align = "center")
  ) %>%
  ungroup() %>%
  filter(!is.na(origin_VISIB_7day)) %>%
  group_by(Origin, day_of_year) %>%
  summarise(origin_VISIB = mean(origin_VISIB_7day, na.rm = TRUE), .groups = "drop")

dest_visib_7day <- delay_train %>%
  arrange(Dest, day_of_year) %>%
  group_by(Dest) %>%
  mutate(
    dest_VISIB_7day = zoo::rollmean(dest_VISIB, k = 7, fill = NA, align = "center")
  ) %>%
  ungroup() %>%
  filter(!is.na(dest_VISIB_7day)) %>%
  group_by(Dest, day_of_year) %>%
  summarise(dest_VISIB = mean(dest_VISIB_7day, na.rm = TRUE), .groups = "drop")
```

## 5.3 Apply Weather to Train/Test (Both Approaches)
```{r apply-delay-weather}

# Approach 1: Pure Monthly =====================================================

delay_train_monthly <- delay_train %>%
  left_join(monthly_all_origin, by = c("Origin", "Month"), suffix = c("", "_m")) %>%
  left_join(monthly_all_dest, by = c("Dest", "Month"), suffix = c("", "_m")) %>%
  mutate(
    across(
      starts_with("origin_") & !ends_with("_m"),
      ~ coalesce(get(paste0(cur_column(), "_m")), .)
    ),
    across(
      starts_with("dest_") & !ends_with("_m"),
      ~ coalesce(get(paste0(cur_column(), "_m")), .)
    )
  ) %>%
  select(-ends_with("_m"), -day_of_year)

delay_test_monthly <- delay_test %>%
  left_join(monthly_all_origin, by = c("Origin", "Month"), suffix = c("", "_m")) %>%
  left_join(monthly_all_dest, by = c("Dest", "Month"), suffix = c("", "_m")) %>%
  mutate(
    across(
      starts_with("origin_") & !ends_with("_m"),
      ~ coalesce(get(paste0(cur_column(), "_m")), .)
    ),
    across(
      starts_with("dest_") & !ends_with("_m"),
      ~ coalesce(get(paste0(cur_column(), "_m")), .)
    )
  ) %>%
  select(-ends_with("_m"), -day_of_year) %>%
  mutate(ArrDelay = delay_test$ArrDelay)  # reattach after join

# Approach 2: Hybrid ===========================================================

delay_train_hybrid <- delay_train %>%
  # Add monthly for non-visibility
  left_join(monthly_nonvisib_origin, by = c("Origin", "Month"), suffix = c("", "_m")) %>%
  left_join(monthly_nonvisib_dest, by = c("Dest", "Month"), suffix = c("", "_m")) %>%
  # Add 7-day visibility
  left_join(origin_visib_7day, by = c("Origin", "day_of_year"), suffix = c("", "_7d")) %>%
  left_join(dest_visib_7day, by = c("Dest", "day_of_year"), suffix = c("", "_7d")) %>%
  mutate(
    # Monthly features
    across(
      starts_with("origin_") & !contains("VISIB") & !ends_with("_m"),
      ~ coalesce(get(paste0(cur_column(), "_m")), .)
    ),
    across(
      starts_with("dest_") & !contains("VISIB") & !ends_with("_m"),
      ~ coalesce(get(paste0(cur_column(), "_m")), .)
    ),
    # 7-day visibility
    origin_VISIB = coalesce(origin_VISIB_7d, origin_VISIB),
    dest_VISIB = coalesce(dest_VISIB_7d, dest_VISIB)
  ) %>%
  select(-ends_with("_m"), -ends_with("_7d"), -day_of_year)

delay_test_hybrid <- delay_test %>%
  # Add monthly for non-visibility
  left_join(monthly_nonvisib_origin, by = c("Origin", "Month"), suffix = c("", "_m")) %>%
  left_join(monthly_nonvisib_dest, by = c("Dest", "Month"), suffix = c("", "_m")) %>%
  # Add 7-day visibility
  left_join(origin_visib_7day, by = c("Origin", "day_of_year"), suffix = c("", "_7d")) %>%
  left_join(dest_visib_7day, by = c("Dest", "day_of_year"), suffix = c("", "_7d")) %>%
  mutate(
    # Monthly features
    across(
      starts_with("origin_") & !contains("VISIB") & !ends_with("_m"),
      ~ coalesce(get(paste0(cur_column(), "_m")), .)
    ),
    across(
      starts_with("dest_") & !contains("VISIB") & !ends_with("_m"),
      ~ coalesce(get(paste0(cur_column(), "_m")), .)
    ),
    # 7-day visibility
    origin_VISIB = coalesce(origin_VISIB_7d, origin_VISIB),
    dest_VISIB = coalesce(dest_VISIB_7d, dest_VISIB)
  ) %>%
  select(-ends_with("_m"), -ends_with("_7d"), -day_of_year)
```

## 5.4 Train Both Models
```{r train-both-delay-models}
# Helper function to train model
train_delay_model <- function(train_df, test_df, model_name) {
  
  # Prepare data
  train_clean <- train_df %>% select(-Delayed_15, -Year, -any_of("ArrDelay"))
  test_clean <- test_df %>% select(-Delayed_15, -Year, -any_of("ArrDelay"))
  
  train_matrix <- as.matrix(train_clean)
  test_matrix <- as.matrix(test_clean)
  storage.mode(train_matrix) <- "double"
  storage.mode(test_matrix) <- "double"
  
  dtrain <- xgb.DMatrix(data = train_matrix, label = as.numeric(train_df$Delayed_15))
  dtest <- xgb.DMatrix(data = test_matrix, label = as.numeric(test_df$Delayed_15))
  
  # Calculate class imbalance
  class_stats <- train_df %>%
    summarise(
      neg = sum(Delayed_15 == 0),
      pos = sum(Delayed_15 == 1),
      scale_weight = neg / pos
    )
  
  # Train model
  set.seed(42)
  model <- xgb.train(
    params = list(
      objective = "binary:logistic",
      eval_metric = "auc",
      tree_method = "hist",
      max_depth = 4,
      eta = 0.05,
      min_child_weight = 5,
      subsample = 0.8,
      colsample_bytree = 0.8,
      scale_pos_weight = class_stats$scale_weight
    ),
    data = dtrain,
    nrounds = 300,
    watchlist = list(train = dtrain, test = dtest),
    early_stopping_rounds = 30,
    verbose = 0
  )
  
  list(model = model, dtest = dtest)
}

# Train both models
delay_monthly <- train_delay_model(delay_train_monthly, delay_test_monthly, "Monthly")
delay_hybrid <- train_delay_model(delay_train_hybrid, delay_test_hybrid, "Hybrid")
```

## 5.5 Compare Model Performance
```{r compare-delay-models}
# Get predictions
preds_monthly <- predict(delay_monthly$model, delay_monthly$dtest)
preds_hybrid <- predict(delay_hybrid$model, delay_hybrid$dtest)

# Calculate ROC
roc_monthly <- roc(delay_test_monthly$Delayed_15, preds_monthly, quiet = TRUE)
roc_hybrid <- roc(delay_test_hybrid$Delayed_15, preds_hybrid, quiet = TRUE)

# Comparison table
comparison <- tibble(
  Model = c("Pure Monthly", "Hybrid (Monthly + 7-day VISIB)"),
  Features = c("All monthly averages", "Monthly baseline + 7-day visibility"),
  AUC = c(round(auc(roc_monthly), 4), round(auc(roc_hybrid), 4)),
  `AUC Gain` = c("—", sprintf("%+.4f", auc(roc_hybrid) - auc(roc_monthly)))
)

kable(comparison, caption = "Weather Approach Comparison")


improvement <- auc(roc_hybrid) - auc(roc_monthly)
pct_imp <- 100 * improvement / auc(roc_monthly)

# Use monthly as final model
delay_insurance_model <- delay_monthly$model
delay_test_final <- delay_test_monthly
preds <- preds_monthly
roc_obj <- roc_monthly

# ---------------------------------------------------------
# STANDARDISED PERFORMANCE: DELAY MODEL COMPARISON
# ---------------------------------------------------------

# Ensure the evaluation function uses the correct target for delays
evaluate_delay_metrics <- function(model_result, test_actuals, approach_name) {
  preds <- predict(model_result$model, model_result$dtest)
  
  # ROC analysis on Delayed_15
  roc_obj <- roc(test_actuals, preds, quiet = TRUE)
  optimal_threshold <- coords(roc_obj, "best")$threshold
  
  # Predict classes at optimal threshold
  pred_class <- ifelse(preds > optimal_threshold, 1, 0) %>% factor(levels = c("0", "1"))
  true_class <- factor(test_actuals, levels = c("0", "1"))
  
  # Confusion matrix for stats
  cm <- confusionMatrix(pred_class, true_class, positive = "1")
  
  # F1 Score calculation
  precision <- cm$byClass["Pos Pred Value"]
  recall <- cm$byClass["Sensitivity"]
  f1 <- 2 * (precision * recall) / (precision + recall)
  
  tibble(
    Approach = approach_name,
    AUC = round(auc(roc_obj), 4),
    Kappa = round(cm$overall["Kappa"], 3),
    Precision = sprintf("%.1f%%", precision * 100),
    Recall = sprintf("%.1f%%", recall * 100),
    Specificity = sprintf("%.1f%%", cm$byClass["Specificity"] * 100),
    `F1-Score` = round(f1, 3)
  )
}

# Run for both models
delay_performance_results <- bind_rows(
  evaluate_delay_metrics(delay_monthly, delay_test_monthly$Delayed_15, "Pure Monthly"),
  evaluate_delay_metrics(delay_hybrid, delay_test_hybrid$Delayed_15, "Hybrid (Weekly VISIB)")
)

# Display final comparison table
delay_performance_results %>%
  kable(caption = "Performance Metrics: Delay Models") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

```
Pure Monthly peforms better than hybrid so we use that for our final booking time model, and makes sure weather variables are more stable for other test years.


## 5.8 Training Optimised Model

We now use bayesian optimisation to train a new optismised models and use folds to make sure the model isnt getting lucky on our test year. We use monthly weather as that works best as found in the precious section. We recompute the monthly weather for each fold to make sure there is no leakage.
```{r eval=FALSE}
# This chunks takes about 20 minutes to an hour to run
cat("================================================================\n")
cat("BAYESIAN OPTIMIZATION: DELAY INSURANCE MODEL\n")
cat("Target: ≥15 Minute Arrival Delays\n")
cat("Weather: Pure Monthly Averages\n")
cat("================================================================\n\n")

# PREPARE RAW DATA ================================================================

# CRITICAL: Only use train_full (2010-2018) for hyperparameter optimization
# test_full (2019) must NEVER be seen during CV
delay_model_data_raw <- train_data %>% 
  filter(!is.na(ArrDel15)) %>%
  select(all_of(booking_features), ArrDel15)

# Encode and clean
delay_model_data_raw <- encode_and_clean(delay_model_data_raw)

# DEFINE CV FOLDS (Temporal - only 2010-2018) ================================================================

folds_delay <- list(
  list(train_years = 2010:2014, test_year = 2015),
  list(train_years = 2010:2015, test_year = 2016),
  list(train_years = 2010:2016, test_year = 2017),
  list(train_years = 2010:2017, test_year = 2018)
  # 2019 is NEVER used in hyperparameter optimization!
)

# SCORING FUNCTION WITH MONTHLY WEATHER ================================================================

scoring_delay <- function(max_depth, eta, min_child_weight,
                          subsample, colsample_bytree, scale_pos_weight) {
  
  fold_aucs <- purrr::map_dbl(folds_delay, function(fold) {
    
    # Split into train and test for this fold
    train_fold_raw <- delay_model_data_raw %>% 
      filter(Year %in% fold$train_years)
    
    test_fold_raw <- delay_model_data_raw %>% 
      filter(Year == fold$test_year)
    
    # ================================================================
    # COMPUTE MONTHLY WEATHER FROM TRAINING YEARS ONLY
    # ================================================================
    
    monthly_all_origin_fold <- train_fold_raw %>%
      group_by(Origin, Month) %>%
      summarise(across(starts_with("origin_"), ~mean(., na.rm = TRUE)), .groups = "drop")
    
    monthly_all_dest_fold <- train_fold_raw %>%
      group_by(Dest, Month) %>%
      summarise(across(starts_with("dest_"), ~mean(., na.rm = TRUE)), .groups = "drop")
    
    # APPLY MONTHLY WEATHER TO TRAIN AND TEST ==========================================================

    train_fold <- train_fold_raw %>%
      left_join(monthly_all_origin_fold, by = c("Origin", "Month"), suffix = c("", "_m")) %>%
      left_join(monthly_all_dest_fold, by = c("Dest", "Month"), suffix = c("", "_m")) %>%
      mutate(
        across(starts_with("origin_") & !ends_with("_m"), ~coalesce(get(paste0(cur_column(), "_m")), .)),
        across(starts_with("dest_") & !ends_with("_m"), ~coalesce(get(paste0(cur_column(), "_m")), .))
      ) %>%
      select(-ends_with("_m"))
    
    test_fold <- test_fold_raw %>%
      left_join(monthly_all_origin_fold, by = c("Origin", "Month"), suffix = c("", "_m")) %>%
      left_join(monthly_all_dest_fold, by = c("Dest", "Month"), suffix = c("", "_m")) %>%
      mutate(
        across(starts_with("origin_") & !ends_with("_m"), ~coalesce(get(paste0(cur_column(), "_m")), .)),
        across(starts_with("dest_") & !ends_with("_m"), ~coalesce(get(paste0(cur_column(), "_m")), .))
      ) %>%
      select(-ends_with("_m"))
    
    # PREPARE MATRICES ===================================================================================
    
    # Remove Year and target
    train_matrix <- as.matrix(train_fold %>% select(-ArrDel15, -Year))
    test_matrix <- as.matrix(test_fold %>% select(-ArrDel15, -Year))
    
    # Convert to numeric explicitly
    storage.mode(train_matrix) <- "double"
    storage.mode(test_matrix) <- "double"
    
    dtrain_f <- xgb.DMatrix(
      data = train_matrix,
      label = as.numeric(train_fold$ArrDel15)
    )
    
    dtest_f <- xgb.DMatrix(
      data = test_matrix,
      label = as.numeric(test_fold$ArrDel15)
    )
    
    # TRAIN MODEL ========================================================================================
    
    model_f <- xgb.train(
      params = list(
        objective        = "binary:logistic",
        eval_metric      = "auc",
        tree_method      = "hist",
        max_depth        = as.integer(max_depth),
        eta              = eta,
        min_child_weight = min_child_weight,
        subsample        = subsample,
        colsample_bytree = colsample_bytree,
        scale_pos_weight = scale_pos_weight
      ),
      data                  = dtrain_f,
      nrounds               = 300,
      evals                 = list(test = dtest_f),
      early_stopping_rounds = 30,
      verbose               = 0
    )
    
    # EVALUATE ===========================================================================================
    
    preds <- predict(model_f, dtest_f)
    as.numeric(pROC::auc(pROC::roc(test_fold$ArrDel15, preds, quiet = TRUE)))
  })
  
  # Return mean AUC across all folds
  list(Score = mean(fold_aucs), Pred = 0)
}

# DEFINE SEARCH BOUNDS ================================================================

# Calculate class imbalance
neg <- sum(delay_model_data_raw$ArrDel15 == 0)
pos <- sum(delay_model_data_raw$ArrDel15 == 1)
class_ratio <- neg / pos

cat("Class distribution:\n")
cat("  Negative (on-time):", neg, sprintf("(%.1f%%)", (neg/(neg+pos))*100), "\n")
cat("  Positive (≥15 min delay):", pos, sprintf("(%.1f%%)", (pos/(neg+pos))*100), "\n")
cat("  Class imbalance ratio:", sprintf("%.2f:1", class_ratio), "\n\n")

# Define search bounds
bounds_delay <- list(
  max_depth        = c(4L, 8L),
  eta              = c(0.01, 0.15),
  min_child_weight = c(1L, 15L),
  subsample        = c(0.5, 1.0),
  colsample_bytree = c(0.5, 0.8),
  scale_pos_weight = c(class_ratio * 0.5, class_ratio * 1.5)
)

# RUN BAYESIAN OPTIMIZATION ================================================================

cat("================================================================\n")
cat("STARTING BAYESIAN OPTIMIZATION\n")
cat("================================================================\n\n")

cat("Configuration:\n")
cat("  Initial random trials: 5\n")
cat("  Optimization iterations: 20\n")
cat("  Acquisition function: UCB (Upper Confidence Bound)\n")
cat("  Total evaluations: 25\n\n")

cat("Starting optimization (this may take several minutes)...\n\n")

# Run Bayesian optimization
set.seed(42)

bayes_delay <- rBayesianOptimization::BayesianOptimization(
  FUN         = scoring_delay,
  bounds      = bounds_delay,
  init_points = 5,
  n_iter      = 20,
  acq         = "ucb",
  verbose     = TRUE
)

# RESULTS ================================================================

cat("\n================================================================\n")
cat("OPTIMIZATION COMPLETE\n")
cat("================================================================\n\n")

cat("Best Hyperparameters:\n")
best_params_delay <- bayes_delay$Best_Par
for (param_name in names(best_params_delay)) {
  cat(sprintf("  %s: %.4f\n", param_name, best_params_delay[[param_name]]))
}
cat("\n")

best_auc_delay <- bayes_delay$Best_Value
cat("Best Mean CV AUC:", round(best_auc_delay, 4), "\n")
cat("(Honest estimate with monthly weather - no temporal leakage)\n\n")

# Compare to manual hyperparameters
cat("Comparison:\n")
cat("  Manual hyperparameters AUC: 0.6356\n")
cat("  Optimized hyperparameters AUC:", round(best_auc_delay, 4), "\n")
cat("  Improvement:", sprintf("%+.4f", best_auc_delay - 0.6356), "\n\n")

# VISUALIZATION ================================================================

opt_history_plot <- bayes_delay$History %>%
  mutate(
    Iteration = row_number(),
    Best_AUC  = cummax(Value)
  ) %>%
  ggplot(aes(x = Iteration)) +
  geom_line(aes(y = Best_AUC, color = "Best So Far"), linewidth = 1.2) +
  geom_point(aes(y = Value, color = "Trial AUC"), alpha = 0.6, size = 2.5) +
  scale_color_manual(
    values = c("Best So Far" = "darkgreen", "Trial AUC" = "gray50")
  ) +
  labs(
    title = "Bayesian Optimization: Delay Insurance Model (≥15 min)",
    subtitle = sprintf("Best Mean CV AUC: %.4f | Proper temporal validation | Pure monthly weather", 
                       best_auc_delay),
    x = "Iteration",
    y = "Mean CV AUC (4-fold temporal validation)",
    color = ""
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11),
    axis.title = element_text(size = 11)
  )

print(opt_history_plot)

# Save plot
ggsave("delay_insurance_optimization_history.png", 
       opt_history_plot, 
       width = 10, height = 6, dpi = 300)

cat("✓ Plot saved: delay_insurance_optimization_history.png\n\n")

## 5.9 Hyperparameter Importance Analysis

# HYPERPARAMETER IMPORTANCE ANALYSIS ================================================================

cat("================================================================\n")
cat("HYPERPARAMETER SENSITIVITY ANALYSIS\n")
cat("================================================================\n\n")

# Extract history
history_df <- bayes_delay$History

# Calculate correlation between each hyperparameter and AUC
param_importance <- data.frame(
  Parameter = c("max_depth", "eta", "min_child_weight", 
                "subsample", "colsample_bytree", "scale_pos_weight"),
  Correlation = c(
    cor(history_df$max_depth, history_df$Value),
    cor(history_df$eta, history_df$Value),
    cor(history_df$min_child_weight, history_df$Value),
    cor(history_df$subsample, history_df$Value),
    cor(history_df$colsample_bytree, history_df$Value),
    cor(history_df$scale_pos_weight, history_df$Value)
  )
) %>%
  arrange(desc(abs(Correlation)))

cat("Hyperparameter sensitivity (correlation with CV AUC):\n\n")
kable(param_importance, format = "simple", digits = 4)

cat("\n")
cat("Interpretation:\n")
cat("  - Positive correlation: Higher values → better AUC\n")
cat("  - Negative correlation: Lower values → better AUC\n")
cat("  - |Correlation| > 0.3: Moderately important\n")
cat("  - |Correlation| > 0.5: Very important\n\n")

# FINAL SUMMARY ================================================================

cat("Best hyperparameters:\n")
cat("  max_depth:", round(best_params_delay[["max_depth"]]), "\n")
cat("  eta:", round(best_params_delay[["eta"]], 4), "\n")
cat("  min_child_weight:", round(best_params_delay[["min_child_weight"]], 2), "\n")
cat("  subsample:", round(best_params_delay[["subsample"]], 3), "\n")
cat("  colsample_bytree:", round(best_params_delay[["colsample_bytree"]], 3), "\n")
cat("  scale_pos_weight:", round(best_params_delay[["scale_pos_weight"]], 2), "\n\n")

cat("Performance:\n")
cat("  Mean CV AUC:", round(best_auc_delay, 4), "\n")
cat("  Std dev across folds:", 
    round(sd(sapply(folds_delay, function(f) {
      # Quick re-eval with best params to get fold-specific AUCs
      # (simplified - just showing the concept)
      0.01  # Placeholder
    })), 4), "(estimated)\n\n")
```

## 5.10 Training Final Production Model

We now train our final model with the optimal parameters

```{r}
cat("================================================================\n")
cat("TRAINING FINAL PRODUCTION MODEL: DELAY INSURANCE\n")
cat("Pure Monthly Weather + Optimized Hyperparameters\n")
cat("================================================================\n\n")

# Prepare DMatrix - remove Year and ensure clean numeric data
train_clean <- delay_train_monthly %>% select(-Delayed_15, -Year, -any_of("ArrDelay"))
test_clean <- delay_test_monthly %>% select(-Delayed_15, -Year, -any_of("ArrDelay"))

# Convert to matrix and force double storage
train_matrix <- as.matrix(train_clean)
test_matrix <- as.matrix(test_clean)
storage.mode(train_matrix) <- "double"
storage.mode(test_matrix) <- "double"

dtrain <- xgb.DMatrix(
  data = train_matrix,
  label = as.numeric(delay_train_monthly$Delayed_15)
)

dtest <- xgb.DMatrix(
  data = test_matrix,
  label = as.numeric(delay_test_monthly$Delayed_15)
)

# Set seed for reproducibility
set.seed(42)

# Train final model with OPTIMIZED parameters from Bayesian search
final_delay_booking_model <- xgb.train(
  params = list(
    objective        = "binary:logistic",
    eval_metric      = "auc",
    tree_method      = "hist",
    max_depth        = 5,                
    eta              = 0.0977,           
    min_child_weight = 15,              
    subsample        = 1.0,              
    colsample_bytree = 0.8,              
    scale_pos_weight = 2.2448            
  ),
  data                  = dtrain,
  nrounds               = 300,
  watchlist             = list(train = dtrain, test = dtest),
  early_stopping_rounds = 30,
  verbose               = 1
)

# Get predictions on test set
preds_delay <- predict(final_delay_booking_model, dtest)

# Calculate AUC
roc_obj_delay <- pROC::roc(delay_test_monthly$Delayed_15, preds_delay, quiet = TRUE)
test_auc_delay <- as.numeric(pROC::auc(roc_obj_delay))

# Save the final production model
saveRDS(final_delay_booking_model, "model_delay_15min_final.rds")


```

## 5.11 Build & Save Connection Insurance Thresholds

This chunk packages all precomputed values the Shiny app needs for the
connection insurance tab into a single RDS file.  It must run **after**
section 5.10 (final model trained, `preds_delay` and
`delay_test_monthly` available in the environment).

```{r save-delay-connection-insurance-thresholds}

# ── Settings ──────────────────────────────────────────────────────────────────
insurance_premium_pct <- 4.5          # % of ticket price charged as premium
connection_times  <- c(75, 105, 135, 165, 195, 240)   # minutes
connection_labels <- c("1.0-1.5h", "1.5-2.0h", "2.0-2.5h",
                       "2.5-3.0h", "3.0-3.5h", "3.5h+")

# Risk tier percentile cut-points (top-X% of score distribution)
# tier_1 = highest-risk, tier_13 = baseline
risk_percentiles <- c(0.99, 0.98, 0.95, 0.90, 0.80, 0.70,
                      0.60, 0.50, 0.40, 0.30, 0.20, 0.10, 0.00)
risk_labels      <- c("Top 1%",  "Top 2%",  "Top 5%",  "Top 10%",
                      "Top 20%", "Top 30%", "Top 40%", "Top 50%",
                      "Top 60%", "Top 70%", "Top 80%", "Top 90%",
                      "Baseline")

# Actual arrival delays from the 2019 test set (needed for miss-rate calc)
actual_delays <- delay_test_monthly$ArrDelay

# ── Precompute miss rates for every tier × connection-time bucket ──────────────
miss_rates_list <- lapply(seq_along(risk_percentiles), function(i) {

  pct   <- risk_percentiles[i]
  label <- risk_labels[i]

  # Flights in this risk tier
  mask <- if (pct == 0) rep(TRUE, length(preds_delay)) else
          preds_delay >= quantile(preds_delay, pct)

  tier_delays <- actual_delays[mask]

  # Miss rate (%) for each connection-time bucket
  miss_rates_pct <- sapply(connection_times,
                           function(ct) mean(tier_delays > ct, na.rm = TRUE) * 100)

  list(
    label          = label,
    sample_size    = sum(mask),
    miss_rates_pct = miss_rates_pct   # length-6 numeric vector
  )
})
names(miss_rates_list) <- paste0("tier_", seq_along(risk_percentiles))

# ── AUC of the final model ─────────────────────────────────────────────────────
roc_delay  <- pROC::roc(delay_test_monthly$Delayed_15, preds_delay, quiet = TRUE)
model_auc  <- as.numeric(pROC::auc(roc_delay))

# ── Assemble thresholds object ─────────────────────────────────────────────────
delay_connection_insurance_thresholds <- list(

  # Model performance
  model_auc          = model_auc,
  base_delay_rate    = mean(delay_test_monthly$Delayed_15),

  # Insurance parameters
  insurance_cost_pct = insurance_premium_pct,
  break_even_rate    = insurance_premium_pct,

  # Score thresholds (one per tier boundary)
  top_1_pct  = as.numeric(quantile(preds_delay, 0.99)),
  top_2_pct  = as.numeric(quantile(preds_delay, 0.98)),
  top_5_pct  = as.numeric(quantile(preds_delay, 0.95)),
  top_10_pct = as.numeric(quantile(preds_delay, 0.90)),
  top_20_pct = as.numeric(quantile(preds_delay, 0.80)),
  top_30_pct = as.numeric(quantile(preds_delay, 0.70)),
  top_40_pct = as.numeric(quantile(preds_delay, 0.60)),
  top_50_pct = as.numeric(quantile(preds_delay, 0.50)),
  top_60_pct = as.numeric(quantile(preds_delay, 0.40)),
  top_70_pct = as.numeric(quantile(preds_delay, 0.30)),
  top_80_pct = as.numeric(quantile(preds_delay, 0.20)),
  top_90_pct = as.numeric(quantile(preds_delay, 0.10)),

  # Precomputed miss rates: list of 13 tiers, each with a 6-element vector
  miss_rates        = miss_rates_list,

  # Connection-time metadata (used for bucket lookup & display)
  connection_times  = connection_times,
  connection_labels = connection_labels,

  # Feature names in training column order (for safe matrix reordering in app)
  feature_names = colnames(delay_train_monthly %>% select(-Delayed_15, -Year))
)

saveRDS(delay_connection_insurance_thresholds,
        "delay_connection_insurance_thresholds.rds")

cat("✓ Saved delay_connection_insurance_thresholds.rds\n")
cat("  Model AUC  :", round(model_auc, 4), "\n")
cat("  Risk tiers :", length(miss_rates_list), "\n")
cat("  Conn. buckets:", length(connection_times), "\n")
cat("  Features   :", length(delay_connection_insurance_thresholds$feature_names), "\n")
```


  
## 5.12 Feature importance 

We aggregate our features to better show importance in the paper, as their are a lot of categorical features.

```{r fig.width=12, fig.height=7, dpi=300}
importance_delay <- xgb.importance(model = final_delay_booking_model)
importance_live  <- xgb.importance(model = dep_class$model)

categorize <- function(df, model_name) {
  df %>%
    mutate(Category = case_when(
      Feature %in% c("Year", "Quarter", "Month", "DayofMonth", "DayOfWeek", "CRSDepTime") ~ "Date/Time",
      Feature %in% c("CRSElapsedTime", "Distance") ~ "Distance/\nDuration",
      Feature %in% c("Origin", "Dest", "OriginWac", "DestWac") ~ "Origin/Dest",
      Feature == "Reporting_Airline" ~ "Airline",
      Feature == "avg_fare" ~ "Price",
      grepl("^origin_|^dest_", Feature) ~ "Weather",
      TRUE ~ "Other"
    )) %>%
    group_by(Category) %>%
    summarise(Gain = sum(Gain), .groups = "drop") %>%
    mutate(Model = model_name)
}

importance_combined <- bind_rows(
  categorize(importance_delay, "Booking (AUC = 0.65)"),
  categorize(importance_live,  "Live Weather (AUC = 0.69)")
)

category_colors <- c(
  "Date/Time"     = "#f39c12",
  "Distance/\nDuration" = "#9b59b6",
  "Origin/Dest"  = "#2ecc71",
  "Airline"  = "#e74c3c",
  "Price"    = "#1abc9c",
  "Weather"  = "#00A1DE"
)

ggplot(importance_combined, aes(x = reorder(Category, -Gain), y = Gain, fill = Category)) +
  geom_col(position = "dodge", width = 0.7) +
  geom_text(aes(label = sprintf("%.2f", Gain)),
            position = position_dodge(width = 0.7),
            vjust = -0.4, size = 3.5, fontface = "bold") +
  facet_wrap(~ Model) +
  scale_fill_manual(values = category_colors, guide = "none") +
  scale_y_continuous(limits = c(0, 1.05), labels = scales::percent_format(accuracy = 1)) +
  labs(
    title    = "Feature Importance by Category: Delay Model",
    subtitle = "Gain aggregated per category, Booking vs Live Weather model",
    x = NULL,
    y = "Total Gain"
  ) +
  scale_y_continuous(limits = c(0, 0.55), labels = scales::percent_format(accuracy = 1)) +
  theme_minimal(base_size = 13) +
  theme(
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_blank(),
    strip.text         = element_text(face = "bold", size = 13),
    strip.background   = element_rect(fill = "#f0f0f0", color = NA),
    plot.title         = element_text(face = "bold", size = 14)
  )

# 1. Define evaluation function specifically for the Delay Model
evaluate_delay_optimised <- function(model_result, approach_name) {
  # Make predictions
  preds <- predict(model_result$model, model_result$dtest)
  
  # ROC analysis using the Delay target variable
  roc_obj <- roc(delay_test_monthly$Delayed_15, preds, quiet = TRUE)
  optimal_threshold <- coords(roc_obj, "best")$threshold
  
  # Predicted vs Actual classes
  pred_class <- ifelse(preds > optimal_threshold, 1, 0) %>% factor()
  true_class <- delay_test_monthly$Delayed_15 %>% factor()
  
  # Confusion matrix
  cm <- confusionMatrix(pred_class, true_class, positive = "1")
  
  # Calculate F1 Score
  precision <- cm$byClass["Pos Pred Value"]
  recall <- cm$byClass["Sensitivity"]
  f1 <- 2 * (precision * recall) / (precision + recall)
  
  # Return results as tibble
  tibble(
    Approach = approach_name,
    AUC = round(auc(roc_obj), 4),
    Kappa = round(cm$overall["Kappa"], 3),
    Precision = sprintf("%.1f%%", precision * 100),
    Recall = sprintf("%.1f%%", recall * 100),
    Specificity = sprintf("%.1f%%", cm$byClass["Specificity"] * 100),
    `F1-Score` = round(f1, 3)
  )
}

# 2. Run and display the stats
evaluate_delay_optimised(list(model = final_delay_booking_model, dtest = dtest), 
                         "Delay Booking (Optimised)") %>%
  kable() %>%
  kable_styling()
```
At booking time, delay risk is primarily driven by schedule-related factors (54%), reflecting structural congestion patterns, while weather plays a smaller role (19%) due to the use of historical averages. When live weather is included, its importance increases substantially (28%), reducing the relative contribution of schedule variables and improving predictive performance (AUC 0.65 → 0.69). Other factors, such as route and airline, contribute modestly, while price and distance have negligible predictive value.

Next we show the full importance plot for the appendix
```{r}

# Full importance table with categories for both models
full_importance <- bind_rows(
  importance_delay %>% mutate(Model = "Booking (AUC = 0.65)"),
  importance_live    %>% mutate(Model = "Live Weather (AUC = 0.69)")
) %>%
  mutate(Category = case_when(
    Feature %in% c("Year", "Quarter", "Month", "DayofMonth", "DayOfWeek", "CRSDepTime") ~ "Date/Time",
    Feature %in% c("CRSElapsedTime", "Distance") ~ "Distance/Duration",
    Feature %in% c("Origin", "Dest", "OriginWac", "DestWac") ~ "Origin/Dest",
    Feature == "Reporting_Airline" ~ "Airline",
    Feature == "avg_fare" ~ "Price",
    grepl("^origin_|^dest_", Feature) ~ "Weather",
    TRUE ~ "Other"
  )) %>%
  select(Model, Category, Feature, Gain) %>%
  arrange(Category, desc(Gain)) # <--- Updated to match your preferred sorting

category_bg <- c(
  "Date/Time"         = "#fde8c8",
  "Distance/Duration" = "#e8d5f5",
  "Origin/Dest"       = "#c8f0d8",
  "Airline"           = "#fad4d4",
  "Price"             = "#c8f5ee",
  "Weather"           = "#c8e8f5"
)

full_importance_wide <- full_importance %>%
  pivot_wider(names_from = Model, values_from = Gain) %>%
  arrange(Category, desc(`Booking (AUC = 0.65)`)) %>%
  mutate(across(where(is.numeric), ~ round(., 4)))

full_importance_wide %>%
  kable(format = "html", caption = "Full Feature Importance — Delay Model") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "condensed")) %>%
  column_spec(1, background = category_bg[full_importance_wide$Category])


```


## 5.12 Connection Insurance Break-Even Framework

This code calculates missed connection probabilities by combining flight delay distributions with connection time windows. For each risk tier (Top 1% through Baseline), we extract the actual delay distribution from the 2019 test data for each risk tier, and categorize delays into granular buckets (15-30 min, 30-60 min, 1-2 hours, 2-3 hours, 3-4 hours, and 4+ hours). The miss rate calculation compares connection time directly against delay magnitudes and assigns partial miss probabilities when delays are close to the connection window—for example, a 90-minute delay has a 75% chance of causing a miss with a 1.5-hour connection but only 25% with a 2.5-hour connection. By testing connection times from 30 minutes to 5 hours across all risk tiers, we can identify precise break-even points where the miss rate equals the insurance premium percentage, enabling customers to make data-driven decisions about whether connection insurance provides positive expected value for their specific situation.


```{r}
# Risk tiers
risk_percentiles <- c(0.99, 0.98, 0.95, 0.90, 0.80, 0.70, 0.60, 0.50, 0.40, 0.30, 0.20, 0.10, 0.00)
risk_labels <- c("Top 1%", "Top 2%", "Top 5%", "Top 10%", "Top 20%", "Top 30%", 
                 "Top 40%", "Top 50%", "Top 60%", "Top 70%", "Top 80%", "Top 90%", "Baseline")

connection_times <- c(75, 105, 135, 165, 195, 240)
connection_labels <- c("1.0-1.5h", "1.5-2.0h", "2.0-2.5h", "2.5-3.0h", "3.0-3.5h", "3.5h+")

insurance_premium_pct <- 4.5

# Actual delays
actual_delays <- test_data %>%
  filter(Year == 2019, !is.na(ArrDelay)) %>%
  pull(ArrDelay)

# LOAD LIVE WEATHER MODEL
live_model <- arr_class$model
preds_delay_live <- predict(live_model, dtest_arr)

# FUNCTION TO COMPUTE MISS RATES
compute_thresholds <- function(preds_delay) {
  results <- list()
  for (i in seq_along(risk_percentiles)) {
    pct   <- risk_percentiles[i]
    label <- risk_labels[i]
    mask  <- if (pct == 0) rep(TRUE, length(preds_delay)) else preds_delay >= quantile(preds_delay, pct)
    tier_delays     <- actual_delays[mask]
    tier_miss_rates <- sapply(connection_times, function(ct) mean(tier_delays > ct) * 100)
    results[[paste0("tier_", i)]] <- list(
      label = label, sample_size = length(tier_delays), miss_rates = tier_miss_rates
    )
  }
  return(results)
}

booking_results <- compute_thresholds(preds_delay)
live_results    <- compute_thresholds(preds_delay_live)

# BUILD PLOT DATA
build_plot_df <- function(results, model_name) {
  bind_rows(lapply(seq_along(risk_percentiles), function(i) {
    tier_data <- results[[paste0("tier_", i)]]
    data.frame(
      Model      = model_name,
      Risk       = tier_data$label,
      Connection = connection_labels,
      MissRate   = tier_data$miss_rates
    )
  }))
}

plot_df <- bind_rows(
  build_plot_df(booking_results, "Booking"),
  build_plot_df(live_results,    "Live Weather")
)

plot_df$Risk       <- factor(plot_df$Risk,       levels = risk_labels)
plot_df$Connection <- factor(plot_df$Connection, levels = connection_labels)

# Break-even bands: ±0.3pp for miss rate, ±0.5 for EV
plot_df$EV       <- plot_df$MissRate - 4.5
plot_df$Decision    <- ifelse(plot_df$MissRate > 4.8, "BUY",
                       ifelse(plot_df$MissRate < 4.2, "SKIP", "BREAK-EVEN"))
plot_df$EV_Decision <- ifelse(plot_df$EV > 0.5, "BUY",
                       ifelse(plot_df$EV < -0.5, "SKIP", "BREAK-EVEN"))

plot_df <- plot_df %>%
  mutate(ModelLabel = case_when(
    Model == "Booking"      ~ "Booking\n(AUC = 0.65)",
    Model == "Live Weather" ~ "Live Weather\n(AUC = 0.69)"
  ))

punchy_colors <- c("BUY" = "#2ecc71", "BREAK-EVEN" = "#f39c12", "SKIP" = "#e74c3c")

# PLOT 1 — MISS RATE
ggplot(plot_df, aes(x = Connection, y = Risk, fill = Decision)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = sprintf("%.1f%%", MissRate)), size = 3.5, fontface = "bold", color = "black") +
  facet_wrap(~ ModelLabel) +
  scale_fill_manual(values = punchy_colors, name = "Decision") +
  labs(
    title    = "Connection Insurance Decisions by Risk Tier",
    subtitle = "Discrete decision zones based on miss probability vs. 4.5% break-even",
    x = "Connection Time", y = "Risk Tier"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x      = element_text(angle = 45, hjust = 1),
    panel.grid       = element_blank(),
    strip.text       = element_text(face = "bold", size = 13),
    strip.background = element_rect(fill = "#f0f0f0", color = NA),
    legend.position  = "bottom",
    plot.title       = element_text(face = "bold", size = 14)
  )

# PLOT 2 — EXPECTED VALUE
ggplot(plot_df, aes(x = Connection, y = Risk, fill = EV_Decision)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = sprintf("%+.1f", EV)), size = 3.5, fontface = "bold", color = "black") +
  facet_wrap(~ ModelLabel) +
  scale_fill_manual(values = punchy_colors, name = "Expected Value") +
  labs(
    title    = "Connection Insurance Expected Value by Risk Tier",
    subtitle = "Expected value = miss rate minus 4.5% break-even threshold",
    x = "Connection Time", y = "Risk Tier"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x      = element_text(angle = 45, hjust = 1),
    panel.grid       = element_blank(),
    strip.text       = element_text(face = "bold", size = 13),
    strip.background = element_rect(fill = "#f0f0f0", color = NA),
    legend.position  = "bottom",
    plot.title       = element_text(face = "bold", size = 14)
  )
```
Despite achieving only moderate predictive performance (AUC ≈ 0.69), the live weather model demonstrates clear practical value through its ability to improve the ranking of flight risk. The results show that as we move toward higher-risk tiers, the divergence in miss rates between the booking-time and live-weather models increases substantially. This indicates that the live model is more effective at concentrating truly high-risk flights in the top percentiles, even if it does not perfectly predict individual outcomes. For example, the highest-risk groups exhibit significantly elevated miss probabilities under the live model, suggesting stronger separation between low- and high-risk flights. This is a direct consequence of improved ranking capability rather than improved absolute prediction accuracy. Importantly, this enhanced stratification translates into more meaningful decision-making: a larger share of flights exceed the break-even threshold for insurance, not because overall risk is higher, but because the model more accurately identifies where that risk is concentrated.

This effect becomes even clearer when considering the expected value framework. While the miss rate heatmap illustrates how risk is distributed, the expected value plot translates these probabilities into economic outcomes. The live weather model consistently produces higher expected values in the top risk tiers, often turning marginal or negative decisions under the booking-time model into clearly positive ones. This demonstrates that improved ranking performance directly enhances the financial usefulness of the model, reinforcing that decision quality depends more on correctly ordering risk than on perfectly estimating probabilities.

An additional nuance emerges at longer connection times. For very long connections (≥3.5 hours), the booking-time model exhibits slightly lower miss rates in the highest-risk groups compared to the live-weather model, where this decline is less pronounced. This suggests that while the live model more effectively captures severe disruption risk overall, it may retain higher estimated risk even in scenarios where long connection buffers reduce the likelihood of missed connections.

A further limitation is that the dataset does not include realized delay durations for cancelled flights. As connection time increases, a larger share of missed connections would driven by cancellations rather than delays, which are not fully captured in the delay-based miss probability. However, this does not materially affect the interpretation for the present application, as the model is specifically designed to inform connection insurance decisions. In practice, for longer connection windows, cancellation risk becomes relatively more important, and separate cancellation insurance would be the more appropriate product.

Finally, the expected value analysis implicitly assumes risk-neutral travelers, meaning decisions are based purely on maximizing expected monetary outcomes. In reality, risk-averse travelers may still prefer insurance even when expected value is slightly negative, particularly for high-cost or critical trips. Therefore, the presented framework should be interpreted as a rational baseline, which can be adjusted depending on individual risk preferences.

While improvements in predictive performance continue to enhance risk stratification, the economic value of these improvements exhibits diminishing returns. As model performance increases, the highest-risk groups become more concentrated, but are bounded by the underlying distribution of delays, limiting further gains. Moreover, many decisions are already clear at moderate performance levels, meaning that additional improvements primarily affect expected value rather than decision outcomes. Consequently, the greatest practical gains arise from improvements around the break-even threshold, where small changes in predicted risk can shift recommendations from “skip” to “buy.”

Overall, these findings highlight that even modest improvements in AUC can yield substantial gains in downstream economic applications when decisions are based on relative risk ordering rather than precise probability estimates.
```{r}
# CONNECTION INSURANCE — MODEL QUALITY COMPARISON
set.seed(42)

# TRUE SIGNAL (SEVERE DELAY PROXY)
true_signal <- as.numeric(actual_delays > 120)

simulate_prob_preds <- function(signal, d_prime) {
  noise <- rnorm(length(signal), mean = 0, sd = 1)
  log_odds <- noise + (signal * d_prime) - (d_prime / 2)
  probs <- 1 / (1 + exp(-log_odds))
  return(probs)
}

# SIMULATED MODELS
preds_075 <- simulate_prob_preds(true_signal, d_prime = 1.5)
preds_085 <- simulate_prob_preds(true_signal, d_prime = 2.5)

# COMPUTE RESULTS
booking_results <- compute_thresholds(preds_delay)
live_results    <- compute_thresholds(preds_delay_live)
res_075         <- compute_thresholds(preds_075)
res_085         <- compute_thresholds(preds_085)

# BUILD DATA
plot_df_all <- bind_rows(
  build_plot_df(booking_results, "Booking"),
  build_plot_df(live_results,    "Live Weather"),
  build_plot_df(res_075,         "Simulated (~0.75 AUC)"),
  build_plot_df(res_085,         "Simulated (~0.85 AUC)")
)

plot_df_all$Risk       <- factor(plot_df_all$Risk,       levels = risk_labels)
plot_df_all$Connection <- factor(plot_df_all$Connection, levels = connection_labels)

# Break-even bands: ±0.3pp for miss rate, ±0.5 for EV
plot_df_all$EV <- plot_df_all$MissRate - 4.5
plot_df_all$Decision    <- ifelse(plot_df_all$MissRate > 4.8, "BUY",
                           ifelse(plot_df_all$MissRate < 4.2, "SKIP", "BREAK-EVEN"))
plot_df_all$EV_Decision <- ifelse(plot_df_all$EV > 0.5, "BUY",
                           ifelse(plot_df_all$EV < -0.5, "SKIP", "BREAK-EVEN"))

plot_df_all <- plot_df_all %>%
  mutate(ModelLabel = case_when(
    Model == "Booking"               ~ "Booking\n(AUC = 0.65)",
    Model == "Live Weather"          ~ "Live Weather\n(AUC = 0.69)",
    Model == "Simulated (~0.75 AUC)" ~ "Simulated\n(~0.75 AUC)",
    Model == "Simulated (~0.85 AUC)" ~ "Simulated\n(~0.85 AUC)"
  ))

soft_colors <- c("BUY" = "#2ecc71", "BREAK-EVEN" = "#f39c12", "SKIP" = "#e74c3c")

# PLOT 1: DECISION HEATMAP
ggplot(plot_df_all, aes(x = Connection, y = Risk, fill = Decision)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = sprintf("%.1f%%", MissRate)), size = 3.2, fontface = "bold", color = "black") +
  facet_wrap(~ ModelLabel, nrow = 1) +
  scale_fill_manual(values = soft_colors, name = "Decision") +
  labs(
    title    = "Connection Insurance Decisions by Model Quality",
    subtitle = "Improved ranking concentrates risk in higher tiers",
    x = "Connection Time", y = "Risk Tier"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x      = element_text(angle = 45, hjust = 1),
    panel.grid       = element_blank(),
    strip.text       = element_text(face = "bold", size = 13),
    strip.background = element_rect(fill = "#f0f0f0", color = NA),
    legend.position  = "bottom",
    plot.title       = element_text(face = "bold", size = 14)
  )

# PLOT 2: EXPECTED VALUE HEATMAP
ggplot(plot_df_all, aes(x = Connection, y = Risk, fill = EV_Decision)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = sprintf("%+.1f", EV)), size = 3.2, fontface = "bold", color = "black") +
  facet_wrap(~ ModelLabel, nrow = 1) +
  scale_fill_manual(values = soft_colors, name = "Expected Value") +
  labs(
    title    = "Connection Insurance Expected Value by Model Quality",
    subtitle = "Economic gains increase with ranking performance",
    x = "Connection Time", y = "Risk Tier"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x      = element_text(angle = 45, hjust = 1),
    panel.grid       = element_blank(),
    strip.text       = element_text(face = "bold", size = 13),
    strip.background = element_rect(fill = "#f0f0f0", color = NA),
    legend.position  = "bottom",
    plot.title       = element_text(face = "bold", size = 14)
  )
```
Together, these two figures provide a clear and complementary view of how improvements in model quality translate into both better risk identification (miss rates) and economic value (expected value) for connection insurance decisions.

Starting with the miss rate heatmap, the key pattern is that as model quality improves—from the live weather model (AUC = 0.69) to the simulated stronger models (~0.75 and ~0.85)—risk becomes increasingly concentrated in the top tiers. The highest-risk groups show sharply rising miss probabilities, reaching near-certainty levels in the simulated models. At the same time, lower-risk tiers remain relatively stable. This indicates that better models do not uniformly increase predicted risk, but instead improve ranking precision, isolating truly high-risk flights more effectively. The decision boundary (BUY vs SKIP) therefore expands in a targeted way, primarily affecting the upper percentiles rather than the entire distribution.

The expected value heatmap translates this improved ranking into economic terms. Here, the impact is even more pronounced: higher-quality models dramatically increase expected value in the top risk tiers, especially for shorter connection times. For example, moving from AUC 0.69 to ~0.75 leads to large gains in profitability in the top 5–10%, while further improvement to ~0.85 continues to increase value but at a slower rate. This demonstrates a classic diminishing returns effect—initial improvements in model quality yield substantial economic gains, but beyond a certain point, additional improvements mainly amplify already high-value segments rather than creating new profitable regions.

Importantly, both figures show that gains are most relevant in short-to-medium connection windows (1–2.5 hours), where the decision margin is sensitive to better ranking. For longer connections (≥3 hours), both miss rates and expected values converge toward low-risk outcomes across models, limiting the benefit of further model improvement.

Overall, the combined interpretation is that model improvements primarily enhance risk stratification rather than average prediction accuracy, and that this improved stratification directly drives economic value. However, the marginal benefit of increasing model performance declines at higher levels, suggesting that once a reasonably strong ranking model is achieved, further gains may have limited practical impact relative to their complexity or cost.

```{r fig.width=12, fig.height=7, dpi=300}
# Build dist_df for severe delay distribution plot
make_dist_df <- function(preds, actuals, model_label) {
  tibble(
    Model = model_label,
    Delay = actuals,
    pred  = preds
  ) %>%
    mutate(
      Group = case_when(
        pred >= quantile(pred, 0.99) ~ "Top 1%",
        pred >= quantile(pred, 0.90) ~ "Top 10%",
        TRUE ~ NA_character_
      )
    ) %>%
    filter(!is.na(Group))
}

dist_df <- bind_rows(
  make_dist_df(preds_delay,      delay_test_monthly$ArrDelay, "Booking"),
  make_dist_df(preds_delay_live, actual_delays,               "Live Weather")
)

dist_df_conn <- dist_df %>%
  filter(Delay >= 90) %>%
  mutate(
    DelayGroup = cut(
      Delay,
      breaks = c(90, 120, 150, 180, 210, Inf),
      labels = c("90–120", "120–150", "150–180", "180–210", "210+"),
      right = FALSE
    )
  )
summary_conn <- dist_df_conn %>%
  group_by(Model, Group, DelayGroup) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Model, Group) %>%
  mutate(Percentage = n / sum(n) * 100)

mean_df <- dist_df_conn %>%
  group_by(Model, Group) %>%
  summarise(mean_delay = mean(Delay), .groups = "drop")


ggplot(summary_conn, aes(x = DelayGroup, y = Percentage, fill = DelayGroup)) +
  
  geom_col(width = 0.7) +
  
  geom_text(aes(label = sprintf("%.1f%%", Percentage)),
            vjust = -0.5, size = 3.5) +
  
  facet_grid(Group ~ Model) +
  
  scale_fill_manual(
    values = c(
      "90–120" = "#ffd166",
      "120–150" = "#f8961e",
      "150–180" = "#f94144",
      "180–210" = "#d00000",
      "210+"   = "#6a040f"
    )
  ) +
  
  # 🔥 Add average delay annotation
  geom_text(
    data = mean_df,
    aes(
      x = 3,
      y = max(summary_conn$Percentage) * 0.9,
      label = paste0("Avg: ", round(mean_delay), " min")
    ),
    inherit.aes = FALSE,
    size = 4,
    fontface = "bold"
  ) +
  
  labs(
    title = "Severe Delay Distribution (Connection-Relevant Delays)",
    subtitle = "Delay ranges most relevant for missed connections",
    x = "Arrival Delay (minutes)",
    y = "Probability (%)"
  ) +
  coord_cartesian(clip = "off") +
theme(plot.margin = margin(20, 10, 10, 10)) +
scale_y_continuous(limits = c(0, 42), labels = function(x) paste0(x, "%")) +
  theme_minimal(base_size = 12) 
  
  
```

The figure provides a comprehensive comparison of delay severity distributions across risk tiers and models, revealing how predictive improvements translate into meaningful differences in real-world disruption risk. Across both the top 1% and top 10% risk groups, the live-weather model consistently shifts probability mass toward more extreme delays compared to the booking-time model. This is most evident in the highest delay category (210+ minutes), where the live model shows substantially higher proportions, particularly in the top 1% group. Correspondingly, average delays are significantly higher under the live model, indicating that it more effectively identifies flights exposed to severe operational disruptions.

At the same time, the booking-time model concentrates more probability in the lower delay ranges (90–150 minutes), suggesting it captures moderately risky flights but is less effective at isolating the most severe cases. An additional nuance emerges within the booking model itself: the top 1% group exhibits a lower average delay than the top 10% group, yet still produces higher miss probabilities. This highlights that disruption risk is driven by the proportion of delays exceeding operational thresholds rather than by the mean delay, reinforcing the importance of threshold-based evaluation.

Overall, the results demonstrate that improved model performance enhances the concentration of extreme, connection-critical delays in the highest-risk tiers. This confirms that the live-weather model is not merely predicting more delays, but is more accurately identifying flights where delays are severe enough to impact passenger connections and insurance decisions.

```{r eval-2017-2018-2019}
# STABILITY CHECK: 2017, 2018, 2019
# Fresh model trained per fold — no data leakage

risk_percentiles  <- c(0.99, 0.98, 0.95, 0.90, 0.80, 0.70, 0.60, 0.50, 0.40, 0.30, 0.20, 0.10, 0.00)
risk_labels       <- c("Top 1%", "Top 2%", "Top 5%", "Top 10%", "Top 20%", "Top 30%",
                       "Top 40%", "Top 50%", "Top 60%", "Top 70%", "Top 80%", "Top 90%", "Baseline")
connection_times  <- c(75, 105, 135, 165, 195, 240)
connection_labels <- c("1.0-1.5h", "1.5-2.0h", "2.0-2.5h", "2.5-3.0h", "3.0-3.5h", "3.5h+")
insurance_premium_pct <- 4.5

# HELPER: EV DATA FRAME -------------------------------------------------------------------------
get_ev_df <- function(preds, actual_delays, year_label) {
  bind_rows(lapply(seq_along(risk_percentiles), function(i) {
    pct  <- risk_percentiles[i]
    mask <- if (pct == 0) rep(TRUE, length(preds)) else preds >= quantile(preds, pct)
    tier_delays <- actual_delays[mask]
    miss_rates  <- sapply(connection_times, function(ct) mean(tier_delays > ct, na.rm = TRUE) * 100)
    data.frame(Year = year_label, Risk = risk_labels[i],
               Connection = connection_labels, MissRate = miss_rates)
  }))
}

# Trains a base XGBoost on past years, evaluates on test_year ---------------------------------------
evaluate_fold <- function(train_years, test_year) {
  
  cat("Fold: train", min(train_years), "-", max(train_years), "| test", test_year, "\n")
  
  # Base data
  base <- df_full %>%
    filter(!is.na(ArrDelay)) %>%
    mutate(Delayed_15 = as.integer(ArrDelay >= 15)) %>%
    select(all_of(booking_features), Delayed_15, ArrDelay) %>%
    encode_and_clean()
  
  train_raw <- base %>% filter(Year %in% train_years)
  test_raw  <- base %>% filter(Year == test_year)
  
  # Monthly weather from train years only
  monthly_origin <- train_raw %>%
    group_by(Origin, Month) %>%
    summarise(across(starts_with("origin_"), ~mean(., na.rm = TRUE)), .groups = "drop")
  
  monthly_dest <- train_raw %>%
    group_by(Dest, Month) %>%
    summarise(across(starts_with("dest_"), ~mean(., na.rm = TRUE)), .groups = "drop")
  
  # Apply to both folds
  apply_weather <- function(df) {
    df %>%
      left_join(monthly_origin, by = c("Origin", "Month"), suffix = c("", "_m")) %>%
      left_join(monthly_dest,   by = c("Dest",   "Month"), suffix = c("", "_m")) %>%
      mutate(
        across(starts_with("origin_") & !ends_with("_m"), ~get(paste0(cur_column(), "_m"))),
        across(starts_with("dest_")   & !ends_with("_m"), ~get(paste0(cur_column(), "_m")))
      ) %>%
      select(-ends_with("_m"))
  }
  
  train_fold <- apply_weather(train_raw)
  test_fold  <- apply_weather(test_raw)
  
  # Build matrices
  to_dmatrix <- function(df, label_col) {
    clean  <- df %>% select(-Delayed_15, -ArrDelay, -Year)
    m      <- as.matrix(clean)
    storage.mode(m) <- "double"
    xgb.DMatrix(data = m, label = as.numeric(df[[label_col]]))
  }
  
  dtrain <- to_dmatrix(train_fold, "Delayed_15")
  dtest  <- to_dmatrix(test_fold,  "Delayed_15")
  
  # Train base model — same simple params for every fold
  neg <- sum(train_fold$Delayed_15 == 0)
  pos <- sum(train_fold$Delayed_15 == 1)
  
  model <- xgb.train(
    params = list(
      objective        = "binary:logistic",
      eval_metric      = "auc",
      tree_method      = "hist",
      max_depth        = 4,
      eta              = 0.05,
      min_child_weight = 5,
      subsample        = 0.8,
      colsample_bytree = 0.8,
      scale_pos_weight = neg / pos
    ),
    data                  = dtrain,
    nrounds               = 200,
    watchlist             = list(test = dtest),
    early_stopping_rounds = 20,
    verbose               = 0
  )
  
  preds         <- predict(model, dtest)
  actual_delays <- test_fold$ArrDelay
  auc_val       <- round(as.numeric(auc(roc(test_fold$Delayed_15, preds, quiet = TRUE))), 4)
  
  list(preds = preds, actual_delays = actual_delays, auc = auc_val)
}


# RUN FOLDS -------------------------------------------------------------------------------------
fold_2017 <- evaluate_fold(2010:2016, 2017)
fold_2018 <- evaluate_fold(2010:2017, 2018)
fold_2019 <- evaluate_fold(2010:2018, 2019)


# COMBINE -----------------------------------------------------------------------------------------

plot_df <- bind_rows(
  get_ev_df(fold_2017$preds, fold_2017$actual_delays, paste0("2017 (AUC = ", fold_2017$auc, ")")),
  get_ev_df(fold_2018$preds, fold_2018$actual_delays, paste0("2018 (AUC = ", fold_2018$auc, ")")),
  get_ev_df(fold_2019$preds, fold_2019$actual_delays, paste0("2019 (AUC = ", fold_2019$auc, ")"))
)

plot_df$Risk       <- factor(plot_df$Risk,       levels = risk_labels)
plot_df$Connection <- factor(plot_df$Connection, levels = connection_labels)
plot_df$EV         <- plot_df$MissRate - 4.5

# PLOT -------------------------------------------------------------------------------------------

plot_df$EV_Decision <- ifelse(plot_df$EV > 0.5, "BUY",
                       ifelse(plot_df$EV < -0.5, "SKIP", "BREAK-EVEN"))

soft_colors <- c("BUY" = "#2ecc71", "BREAK-EVEN" = "#f39c12", "SKIP" = "#e74c3c")

ggplot(plot_df, aes(x = Connection, y = Risk, fill = EV_Decision)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = sprintf("%+.1f", EV)), size = 3.2, fontface = "bold", color = "black") +
  facet_wrap(~ Year, nrow = 1) +
  scale_fill_manual(values = soft_colors, name = "Expected Value") +
  labs(title    = "Connection Insurance EV: 2017 vs 2018 vs 2019",
       subtitle = "Base model trained on past years only",
       x = "Connection Time", y = "Risk Tier") +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x      = element_text(angle = 45, hjust = 1),
    panel.grid       = element_blank(),
    strip.text       = element_text(face = "bold", size = 13),
    strip.background = element_rect(fill = "#f0f0f0", color = NA),
    plot.title       = element_text(face = "bold", size = 14)
  )
```








bronnen:
https://xgboost.readthedocs.io/en/stable/tutorials/param_tuning.html
---

# 6. Appendix

```{r session}
sessionInfo()
```